---
title: "Prices drivers"
author: "Cameron Roach"
date: "23 March 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

#dataDir <- "./docs/histPrice"
dataDir <- "."
#genDataFile <- "./data/HistData/gen_PT.csv"
genDataFile <- "./../../data/HistData/gen_PT.csv"

require(dplyr)
require(lubridate)
require(tidyr)
require(ggplot2)

prices <- read.csv(file.path(dataDir, "histPrice.csv")) %>% 
  rename(Date = DATA,
         Session = SESSAO,
         Hour = HORA,
         Price_PT = PRECO_PT,
         Price_ES = PRECO_ES) %>% 
  mutate(ts = dmy(Date) + hours(Hour-1)) %>% 
  filter(Session==0) %>% 
  select(-c(Hour, Session, Price_ES, Date))

gen <- read.csv(genDataFile) %>% 
  mutate(ts = ymd_hms(date)) %>% 
  select(-c(date, id))

df <- inner_join(prices, gen)
```

## Historical price drop

There appears to be a level shift between 2015 and 2016 in prices.

```{r}
prices %>% 
  ggplot(aes(x=ts, y=Price_PT)) +
  geom_line() +
  ggtitle("Portugal electricity prices.")
```

### Generation

This may be caused by generation type. Here are the generation time series for Portugal.

```{r}
df %>% 
  gather(Generation, MWh, -c(ts, Price_PT)) %>%
  ggplot(aes(x=ts, y=Price_PT)) +
  geom_line() +
  facet_wrap(~Generation) +
  ggtitle("Generation time series.")
```

Unfortunately, the generation data only extends to 31 December 2015 and so the change in relationship after the level shift occurs can't really be observed.

Correlation is shown below.

```{r}
df %>% 
  gather(Generation, MWh, -c(ts, Price_PT)) %>%
  ggplot(aes(x=MWh, y=Price_PT)) +
  geom_point() +
  facet_wrap(~Generation, scales="free_x") +
  geom_smooth() + 
  ggtitle("Correlation generation and price")

```


Remove fuel\_oil and ssg\_wave because no correlation present.

```{r}
df <- select(df, -c(fuel_oil, ssg_wave))

df %>% 
  gather(Generation, MWh, -c(ts, Price_PT)) %>%
  ggplot(aes(x=MWh, y=Price_PT)) +
  geom_point() +
  facet_wrap(~Generation, scales="free_x") +
  geom_smooth() + 
  ggtitle("Correlation generation and price")


```


### Weather